Crunching Large Graphs with Commodity Processors

### 2011-04-11 doc/hotpar2011/softxmt-hotpar2011.tex
post draft submission comments and typo fixes

%% move stuff to defs file
\newcommand{\todo}[1]{{\bfseries [[#1]]}}
%% To disable, just uncomment this line
 \renewcommand{\todo}[1]{\relax}

@@

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. They do not fit in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near every
other vertex. Their power law distribution means that a few vertices
generate much more work than the rest, leading to load
imbalances. Together, these properties lead to a
difficult conclusion: each edge traversal is likely to lead to a very
expensive inter-node communication.

@@

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent. The goal of our system is to support
that concurrency and consequently provide latency tolerance on a large
address space. We do so via software multithreading: computation that
is about to execute a long latency operation (e.g. a remote memory
reference) initiates the operation and then yields to the scheduler,
which quickly resumes execution of another computation.  Concurrency
is also used---via aggregation of requests to a common node
destination---to make more efficient use of network bandwidth and in
some cases reduce bandwidth demands.

@@

\item[Low overhead synchronization:] 
%  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
  A thread that is synchronizing must yield its core to allow others to make progress.
\end{description}
We discuss our approach to solving each of these challenges in
turn. For each one, we discuss both our current single-node
implementation and ideas for a multi-node implementation.

@@

%We minimize context size by treating context switches as function
%calls, as in~\cite{charm}. This allows us 
%to save and restore the
%minimum number of registers allowed by the ABI, and 
%Simon:  not really true -- we could save fewer if we knew which of the
%callee save registers weren't going to be clobbered.
We treat context switches as function calls, as in~\cite{charm}. This allows us 
to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers.  We minimize
context switch time by doing all switching and scheduling in user space.

@@

A read operation in user code consists of three steps: a {\em read
  issue}, which causes the data to start moving; a {\em yield}, which
causes another thread to start executing, overlapping the read
latency; and, once the reading thread has been rescheduled, a {\em
  read complete} blocks until the data is available. A write operation
follow the same pattern.

@@

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions. We
also observe many points that do not follow the trend are below the max, but these results are repeatable. This effect is harder to explain, but we suspect it is
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

We presented our plans for building a system for large graph
computation using commodity parts. We leverage decades-old ideas on
using a large amount of concurrency to tolerate latencies; but we do so
mostly in software. We developed a runtime system for latency
tolerance and our results showed that it is able to manage enough
concurrency to tolerate long latencies to amortize the cost of
expensive accesses to our large global memory system (high-latency,
but high-bandwidth).

>>>>

%% move stuff to defs file

%command inserts todo/comment
\newcommand{\todo}[1]{{\bfseries [[#1]]}}
%% To disable, just uncomment this line
% \renewcommand{\todo}[1]{\relax}

@@

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. They do not fit in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near every
other vertex. Their power law distribution means that a few vertices
generate much more work than the rest, leading to load
imbalances. Together, these properties \todo{including power law distr?} lead to a
difficult conclusion: each edge traversal is likely to lead to a very
expensive inter-node communication.

@@

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent. The goal of our system is to support
that concurrency and consequently provide latency tolerance on a large
address space. We do so via software multithreading: computation that
is about to execute a long latency operation (e.g. a remote memory
reference) initiates the operation and then yields to the scheduler,
which quickly resumes execution of another computation.  Concurrency
is also used\todo{vague}---via aggregation of requests to a common node
destination---to make more efficient use of network bandwidth and in
some cases reduce bandwidth demands.

@@

\item[Low overhead synchronization:] 
%  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
  A thread that is synchronizing must yield its core to allow others to make progress. \todo{repetitive with the previous item}
\end{description}
We discuss our approach to solving each of these challenges in
turn. For each one, we discuss both our current single-node
implementation and ideas for a multi-node implementation.

@@

%We minimize context size by treating context switches as function
%calls, as in~\cite{charm}. This allows us 
%to save and restore the
%minimum number of registers allowed by the ABI, and 
%Simon:  not really true -- we could save fewer if we knew which of the
%callee save registers weren't going to be clobbered.
\todo{decide whether to change these 2 paragraphs to be inclusive of other lightweight implementations, like jump table}
We treat context switches as function calls, as in~\cite{charm}. This allows us 
to depend on the compiler to save and restore caller-save registers while we explicitly save and restore all the callee-save registers.  We minimize
context switch time by doing all switching and scheduling in user space.

@@

A read operation in user code consists of three steps: a {\em read
  issue}, which causes the data to start moving; a {\em yield}, which
causes another thread to start executing, overlapping the read
latency; and, once the reading thread has been rescheduled, a {\em
  read complete} blocks until the data is available. A write operation
follows the same pattern.

@@

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In the real runtime we expect this effect will not exist, since prefetch will not be implemented with built-in non-temporal prefetch instructions. We
also observe many points that do not follow the trend are below the max, and these results are repeatable. This effect is harder to explain, but we suspect it is
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

We presented our plans for building a system for large graph
computation using commodity parts. We leverage decades-old ideas on
using a large amount of concurrency to tolerate latencies; but we do so
mostly in software. We developed a runtime system for latency
tolerance and our results showed that it is able to manage enough
concurrency to tolerate long latencies to hide the cost of
expensive accesses to our large global memory system (high-latency,
but high-bandwidth).
